%
% file: dedication.tex
% author: V?ctor Bre?a-Medina
% description: Contains the text for thesis dedication
%

\chapter*{Dedication and acknowledgements}
\begin{SingleSpace}
\initial{A} Ph.D. thesis does not happen without external influence. I was incredibly lucky to be given the funding to pursue a project of my own choice. For this I thank the University of Bristol. I would also like to extend my thanks to the invisible hand of the BCCS academic board, whose example has been highly informative. I have learned many things about university life during my time as a postgraduate. I would genuinely like to extend my thanks to EPSRC, whose funding has supported my work since 2011. I sincerely hope that the future will bring with it an enhanced appreciation of the value of public investment in research and education. Not only in the sciences, but in all spheres of knowledge. It is the only way that we can further our true understanding without the pressures of commercial viability. This thesis is an adequate example of the point in question. There is no short term profit to be made from an enhanced understanding of theoretical ecology. Only the catastrophic biodiversity crisis threatening our planet secures adequate funding for the life sciences. I can only hope that the current mental health crisis will soon secure funding for other disciplines that do not appear immediately profitable.       

This thesis would not have been possible without the support of collaborators, advisers, friends and family. I would like to thank them all, but to name a few explicitly. Clive Bowsher was an exemplary supervisor, but unfortunately ours ways parted. Michael Gastner and Lars Rudolph helped me in many ways in the early stages on the project. Daniel Montoya has been an inspiration throughout, and has taught me all that I know about the field of Ecology. Alan Champneys took me on as a student when times were rough, and from him I have learned many things about the art of science. I also thank my collaborators in the realm of public engagement. Specifically, Espen Knoop and Lewis Roberts and all the others involved in the Tangible Networks project. My heartfelt thanks extend to my friends, who have helped me throughout this process, and to my parents whose boundless support I value beyond measure. The ultimate thanks goes to my Kitten. She has experienced the turbulence of this process first hand, and without her I would be but a shallow puddle on the wayside of academia. 

It only remains for me to extend my true thanks to Michael Gove. Without his passion for data driven learning I may never have embarked upon this project.  
\end{SingleSpace}
\clearpage